# Exploration and Summary of White Wine Quality Data \
by Timea Roik

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(psych)
library(corrr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
white_wine_data <- read.csv("wineQualityWhites.csv")

#remove X variable because not relevant information (row number)
white_wine_data$X <- NULL
```

## About the dataset 

This report explores the dataset of white wines and investigates how different 
chemical features may influence the quality ratings.  
The dataset used in this analysis is referenced at the end of the report.  

The white wines samples are from the Portuguese “Vinho Verde” wine.  
For more details, consult: http://www.vinhoverde.pt/en/ or the reference [Cortez 
et al., 2009]. Due to privacy and logistic issues, only physicochemical (inputs) 
and sensory (the output) variables are available (e.g. there is no data about 
grape types, wine brand, wine selling price, etc.).


## Univariate Plots Section

The number of observations and variables in the white wine dataset:

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(white_wine_data)
```

There are 4898 observations and 12 variables.
The main properties of the different variables, including data types and 
values:

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(white_wine_data)
```

Here is a detailed summary of all of the 12 variables:

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(white_wine_data)
```

Since we are interested in the quality ratings of different white wines and 
how certain factors affect those ratings, we need to take a look at
the distribution of the ratings.  The quality rating is a number from 0 to 10, 
with 0 being the worst and 10 being the best quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#999FF9', binwidth = 0.5) +
  stat_bin(binwidth = 1, geom = "text", 
           aes(label = ifelse(..count..> 0, ..count.., "")), 
           vjust = -0.5) +
  scale_x_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) 
```

```{r echo=FALSE}
summary(white_wine_data$quality)
```

Note that the average value for quality is 5.878, but there were no ratings 
given below 3 and above 9. The frequency histogram above shows that the
distribution is close to normal; therefore, the dataset seems to conform to 
the IID requirements.

Then, we can plot the distribution of different factors to have a general 
understanding of the values and outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
ggplot(aes(x = fixed.acidity), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9') +
  scale_x_continuous(limits = c(3.8, 11), breaks = seq(3.8, 11, 2)) 
``` 

```{r echo=FALSE}
summary(white_wine_data$fixed.acidity)
```

Based on the histogram, the values for fixed acidity are distributed normally.
Note that the values for mean and median are very close to each other.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9') +
  scale_x_continuous(limits = c(0.08, 0.78), breaks = seq(0.08, 0.78, 0.2))
```

```{r echo=FALSE}
summary(white_wine_data$volatile.acidity)
```

Based on the histogram, the values for volatile acidity are positively skewed.
Therefore, we will investigate the values after transforming the variable to log
function with base 10.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity), data = white_wine_data) +
  geom_histogram(color = 'black', fill = 'red') +
  scale_x_log10()
```

The distribution is normal if we look at the log values for volatile acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9', binwidth = 0.01) +
  scale_x_continuous(limits = c(0, 0.8), breaks = seq(0, 0.8, 0.2)) 
```

```{r echo=FALSE}
summary(white_wine_data$citric.acid)
```

A few outliers were removed. Based on the histogram, the values for citric acid
are distributed normally. There is one value for citric acid around 0.5 which
has a higher count, increasing the value for the mean.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9', binwidth = 0.3) +
  scale_x_continuous(limits = c(0.6, 20.6), breaks = seq(0.6, 20.6, 5))
```

```{r echo=FALSE}
summary(white_wine_data$residual.sugar)
```

Outliers were removed.
Residual sugar histogram is skewed to the right, so we take a look at the
log value with base 10 for that variable.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar), data = white_wine_data) +
  geom_histogram(color = 'black', fill = 'red') +
  scale_x_log10()
```

This visualization identifies bi-modal character of the distribution with peaks 
around 1.5 and 8 g/dm^3.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = chlorides), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9') +
  scale_x_continuous(limits = c(0.009, 0.1), breaks = seq(0.009, 0.1, 0.025))
```

```{r echo=FALSE}
summary(white_wine_data$chlorides)
```

Outliers were removed.
Based on the histogram, the values for chlorides are normally distributed. The
mean and the median are very close.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = free.sulfur.dioxide), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9') +
  scale_x_continuous(limits = c(2, 120), breaks = seq(2, 120, 35))
```

```{r echo=FALSE}
summary(white_wine_data$free.sulfur.dioxide)
```

From the histogram, it seems that the values for free sulfur dioxide are also
distributed normally.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = total.sulfur.dioxide), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9', binwidth = 6) +
  scale_x_continuous(limits = c(9, 339), breaks = seq(9, 339, 55))
```

```{r echo=FALSE}
summary(white_wine_data$total.sulfur.dioxide)
```

From the histogram, it seems that the values for total sulfur dioxide are also
distributed normally.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = density), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9', binwidth = 0.00035) +
  scale_x_continuous(limits = c(0.9871, 1.0071), 
                     breaks = seq(0.9871, 1.0071, 0.01))
```

```{r echo=FALSE}
summary(white_wine_data$density)
```

From the histogram, it seems that the values for density are distributed 
normally. The mean and the median are very close to each other.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = pH), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9', binwidth = 0.02) +
  scale_x_continuous(limits = c(2.72, 3.82), breaks = seq(2.72, 3.82, 0.25))
```

```{r echo=FALSE}
summary(white_wine_data$pH)
```

From the histogram, it seems that the values for pH are distributed 
normally. The mean and the median are the same.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9', binwidth = 0.11) +
  scale_x_continuous(limits = c(8, 14.2), breaks = seq(8, 14.2, 1))
```

```{r echo=FALSE}
summary(white_wine_data$alcohol)
```

From the histogram, it is not clear whether the values are normally distibutes, 
so I would again plot the log10 histogram for the values of alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol), data = white_wine_data) +
  geom_histogram(color = 'black', fill = 'red', binwidth = 0.05) +
  scale_x_sqrt()
```

After using the square root values for alcohol, the skewness is reduced.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = sulphates), data = white_wine_data) +
  geom_histogram(color = 'black', fill = '#099DD9', binwidth = 0.01) +
  scale_x_continuous(limits = c(0.22, 1.08), breaks = seq(0.22, 1.08, 0.2))
```

```{r echo=FALSE}
summary(white_wine_data$sulphates)
```

From the histogram, it seems that the values for alcohol have a slight skewness 
to the right. Therefore, I would plot the log10 values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = sulphates), data = white_wine_data) +
  geom_histogram(color = 'black', fill = 'red') +
  scale_x_log10()
```

This log10 histogram of sulphates would better represent the normal distribution
of the variables.

## Univariate Analysis

#### What is the structure of your dataset?
There are 4898 observations of white wine data with 12 variables.

Description of attributes, from the original dataset description (see in
References below):  
1. fixed acidity: most acids involved with wine or fixed or nonvolatile 
(do not evaporate readily)  
2. volatile acidity: the amount of acetic acid in wine, which at too high of 
levels can lead to an unpleasant, vinegar taste  
3. citric acid: found in small quantities, citric acid can add ‘freshness’ 
and flavor to wines  
4. residual sugar: the amount of sugar remaining after fermentation stops, 
it’s rare to find wines with less than 1 gram/liter and wines with greater 
than 45 grams/liter are considered sweet  
5. chlorides: the amount of salt in the wine  
6. free sulfur dioxide: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine  
7. total sulfur dioxide: amount of free and bound forms of SO2; in low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine  
8. density: the density of water is close to that of water depending on the 
percent alcohol and sugar content  
9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
to 14 (very basic); most wines are between 3-4 on the pH scale  
10. sulphates: a wine additive which can contribute to sulfur dioxide gas (SO2) 
levels, wich acts as an antimicrobial and antioxidant  
11. alcohol: the percent alcohol content of the wine  

Output variable based on sensory data:   
12. quality (score between 0 and 10)

#### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in the dataset is the quality rating of the
white wines in our sample. We would like to know what influences that score, 
and how.

#### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
The 11 attribute variables from the dataset (listed above) will help support
this investigation in order to find out which attributes have an effect on wine 
quality ratings, and if those effects are positive or negative. 

#### Did you create any new variables from existing variables in the dataset?
No other variables were created from existing variables.

#### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
The variable named 'X' that counted the samples was removed from the dataset, as 
it would not provide any useful information for this analysis. As for the 
unusual distributions, the residual sugar and alcohol contents were not normally 
distributed and had a positive skewness. 

## Bivariate Plots Section

Let's first take a look at a bivariate correlation matrix, to better understand 
how the variables in this dataset are correlated to each other. The closer the 
value is to positive or negative 1, the stronger it is the relationship between 
the variables. It can be positive or negative, depending on the direction of 
change.

```{r echo=FALSE, Bivariate_Plots}
pairs.panels(white_wine_data)
```

```{r echo=FALSE}
x <- white_wine_data %>% 
  correlate() %>% 
  focus(quality)

x
```

```{r echo=FALSE}
x %>% 
  mutate(rowname = factor(rowname, levels = rowname[order(quality)])) %>%  
  # Ordering by correlation strength
  ggplot(aes(x = rowname, y = quality)) +
    geom_bar(stat = "identity") +
    ylab("Correlation with Quality Ratings") +
    xlab("Variable") +
    theme(axis.text.x  = element_text(angle = 90, vjust = 0.5, size = 12))
```

Based on these plots, quality is affected mostly by alcohol content, 
density, and the level of chlorides. As we are most interested in the changes in 
quality, the following graphs present the strongest correlations with the 
quality varible. The relationship between other variables can be seen on the 
bivariate matrix, and will be analyzed in the next section of this report.  

##### Moderate to Low Correlation:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = quality), data = white_wine_data) +
  geom_point(alpha = 1/4, position = 'jitter') +
  geom_smooth(method = 'lm', color = 'red') +
  ggtitle("Quality vs. Alcohol")
```

The red line on the scatterplot confirms that there is indeed a positive
relationship between these two variables. Let's take a look at the boxplots,
grouped by the quality ratings.

```{r echo=FALSE}
ggplot(aes(x = quality, y = alcohol), data = white_wine_data) +
  geom_boxplot(aes(group = quality, fill = factor(quality)), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 10, 1))
```

There are not many outliers for alcohol, although quality rating 5 has a couple 
of outliers. These we not removed. The boxplots illustrate the distribution of 
values for alcohol level for different ratings of quality. We can also take a 
note of the means and how there is a positive trend for quality rating 5 and up.

Correlation coefficient:

```{r echo=FALSE}
cor(white_wine_data$quality, white_wine_data$alcohol)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = density, y = quality), data = white_wine_data) +
  geom_point(alpha = 1/4, position = 'jitter') +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(0.985, 1.005) +
  ggtitle("Quality vs. Density")
```

The red line on the scatterplot confirms that there is indeed a negative
relationship between these two variables. Let's take a look at the boxplots,
grouped by the quality ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = density), data = white_wine_data) +
  geom_boxplot(aes(group = quality, fill = factor(quality)), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  ylim(min(white_wine_data$density), 1.005)
```

There were 2 outliers for density, removed from quality rating 6. Overall, 
there were not many outliers. The boxplots illustrate the distribution of 
values for density for different ratings of quality. We can also take a 
note of the means and how there is a negative trend for quality rating 5 and up.

Correlation coefficient:

```{r echo=FALSE}
cor(white_wine_data$quality, white_wine_data$density)
```

##### Low to Weaker Correlation:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = chlorides, y = quality), data = white_wine_data) +
  geom_point(alpha = 1/4, position = 'jitter') +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(0, 0.1) +
  ggtitle("Quality vs. Chlorides")
```

This scatterplot illustrates low to weaker correlation, between quality and
chlorides with a red line confirming that there is a negative
relationship.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = chlorides), data = white_wine_data) +
  geom_boxplot(aes(group = quality, fill = factor(quality)), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  ylim(min(white_wine_data$chlorides), 0.1) 
```

There were outliers for the quality ratings with the exception of quality 9 that 
had to be removed in order to make the graph more readable. The boxplots 
illustrate the distribution of values for chlorides for different ratings of 
quality. We can also take a note of the means and how there is a negative trend 
for quality rating 4 and up. 

Correlation coefficient:

```{r echo=FALSE}
cor(white_wine_data$quality, white_wine_data$chlorides)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, y = quality), data = white_wine_data) +
  geom_point(alpha = 1/4, position = 'jitter') +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(0.05, 0.75) +
  ggtitle("Quality vs. Volatile Acidity")
```

This scatterplot illustrates low to weaker correlation, between quality and
volatile acidity with a red line confirming that there is a negative
relationship.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = volatile.acidity), data = white_wine_data) +
  geom_boxplot(aes(group = quality, fill = factor(quality)), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  ylim(min(white_wine_data$volatile.acidity), 0.75)
```

There were outliers that had to be removed in order to make the graph more 
readable. The boxplots illustrate the distribution of values for volatile 
acidity levels for different ratings of quality. We can also take a note of the 
means to notice a slight negative trend from quality 4 to 7.

Correlation coefficient:

```{r echo=FALSE}
cor(white_wine_data$quality, white_wine_data$volatile.acidity)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = total.sulfur.dioxide, y = quality), data = white_wine_data) +
  geom_point(alpha = 1/4, position = 'jitter') +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(0, 350) +
  ggtitle("Quality vs. Total Sulfur Dioxide")
```

This scatterplot illustrates low to weaker correlation, between quality and
total sulfur dioxide with a red line confirming that there is a negative
relationship.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = total.sulfur.dioxide), data = white_wine_data) +
  geom_boxplot(aes(group = quality, fill = factor(quality)), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  ylim(min(white_wine_data$volatile.acidity), 350)
```

Outliers that had to be removed again for this boxplot, in order to make it more 
readable. The boxplots illustrate the distribution of values for total sulfur
dioxide levels for different ratings of quality. We can also take a note of the 
means to notice a slight negative trend from quality 5 to 9.

Correlation coefficient:

```{r echo=FALSE}
cor(white_wine_data$quality, white_wine_data$total.sulfur.dioxide)
```

As for the other variables, there are strong correlations between density and
residual sugar, and density and alcohol. These will be also plotted for
reference. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = density), data = white_wine_data) +
  geom_point(alpha = 1/4, position = 'jitter') +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(min(white_wine_data$residual.sugar), 20) +
  ylim(min(white_wine_data$density), 1.0025) +
  ggtitle("Density vs. Residual Sugar")
```

Correlation coefficient:

```{r echo=FALSE}
cor(white_wine_data$density, white_wine_data$residual.sugar)
```

As we can see from the graph, the relationship is very strong. Density and
and residual sugar are positively correlated.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = density, y = alcohol), data = white_wine_data) +
  geom_point(alpha = 1/4, position = 'jitter') +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(min(white_wine_data$density), 1.005) +
  ggtitle("Density vs. Alcohol")
```

Correlation coefficient:

```{r echo=FALSE}
cor(white_wine_data$alcohol, white_wine_data$density)
```

This strong negative relationship is also confirmed by the graph.
Please see the other correlation coefficients between variables and their
interpretations in the analysis section, without graphs.

## Bivariate Analysis

#### Talk about some of the relationships you observed in this part of the investigation. \
How did the feature(s) of interest vary with other features in the dataset?

When looking at the relationship between quality and other variables, we can see
moderate to low correlation between quality and alcohol (+0.44), and quality
and density (-0.31).  
Low to weaker correlation can be found between quality and chlorides (-0.21), 
quality and volatile acidity (-0.19), and quality and total
sulfur dioxide (-0.17).  

#### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

From the bivariate matrix plot, we can also note the following relationships
between other variables. These will be plotted in the next section 
(multivariate plots) in order to see how they affect quality.

Strong Correlation:  
+0.84 density vs. residual sugar  
-0.78 alcohol vs. density  

Moderate Correlation:  
+0.62 free sulfur dioxide vs. total sulfur dioxide  
+0.53 total sulfur dioxide vs. density  

Moderate to Low Correlation:  
-0.45 total sulfur dioxide vs. alcohol  
-0.45 residual sugar vs. alcohol  
-0.43 pH vs. fixed acidity  
+0.40 residual sugar vs. total sulfur dioxide  
-0.36 alcohol vs. chlorides  
+0.30 free sulfur dioxide vs. residual sugar

#### What was the strongest relationship you found?
The strongest relationship is between density and residual sugar content with a
positive correlation of 0.84.

## Multivariate Plots Section

#### Strong Correlation 

Residual Sugar vs. Density By Quality Rating: 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(x = density, y = residual.sugar), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$residual.sugar), 
       quantile(white_wine_data$residual.sugar, 0.95)) +
  xlim(min(white_wine_data$density), quantile(white_wine_data$density, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

Density vs. Alcohol By Quality Rating: 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = density), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$density), 
       quantile(white_wine_data$density, 0.95)) +
  xlim(min(white_wine_data$alcohol), quantile(white_wine_data$alcohol, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

#### Moderate Correlation

Total Sulfur Dioxide vs. Free Sulfur Dioxide By Quality Rating:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$total.sulfur.dioxide), 
       quantile(white_wine_data$total.sulfur.dioxide, 0.95)) +
  xlim(min(white_wine_data$free.sulfur.dioxide), 
       quantile(white_wine_data$free.sulfur.dioxide, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

Total Sulfur Dioxide vs. Density By Quality Rating:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = density, y = total.sulfur.dioxide), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$total.sulfur.dioxide), 
       quantile(white_wine_data$total.sulfur.dioxide, 0.95)) +
  xlim(min(white_wine_data$density), 
       quantile(white_wine_data$density, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

#### Moderate to Low Correlation

Total Sulfur Dioxide vs. Alcohol By Quality Rating:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = total.sulfur.dioxide), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$total.sulfur.dioxide), 
       quantile(white_wine_data$total.sulfur.dioxide, 0.95)) +
  xlim(min(white_wine_data$alcohol), 
       quantile(white_wine_data$alcohol, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

Residual Sugar vs. Alcohol By Quality Rating:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = alcohol), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$alcohol), 
       quantile(white_wine_data$alcohol, 0.95)) +
  xlim(min(white_wine_data$residual.sugar), 
       quantile(white_wine_data$residual.sugar, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

pH vs. Fixed Acidity By Quality Rating:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = pH, y = fixed.acidity), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$fixed.acidity), 
       quantile(white_wine_data$fixed.acidity, 0.95)) +
  xlim(min(white_wine_data$pH), 
       quantile(white_wine_data$pH, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

Residual Sugar vs. Total Sulfur Dioxide By Quality Rating:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = total.sulfur.dioxide), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$total.sulfur.dioxide), 
       quantile(white_wine_data$total.sulfur.dioxide, 0.95)) +
  xlim(min(white_wine_data$residual.sugar), 
       quantile(white_wine_data$residual.sugar, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

Alcohol vs. Chlorides By Quality Rating:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = chlorides), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$chlorides), 
       quantile(white_wine_data$chlorides, 0.95)) +
  xlim(min(white_wine_data$alcohol), 
       quantile(white_wine_data$alcohol, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

Free Sulfur Dioxide vs. Residual Sugar By Quality Rating:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = free.sulfur.dioxide, y = residual.sugar), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$residual.sugar), 
       quantile(white_wine_data$residual.sugar, 0.95)) +
  xlim(min(white_wine_data$free.sulfur.dioxide), 
       quantile(white_wine_data$free.sulfur.dioxide, 0.95)) +
  scale_color_brewer(type = 'seq', palette = 'OrRd')
```

## Multivariate Analysis

#### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

There is a strong correlation between residual sugar and density, and alcohol
and density so the graphs featuring these two variables show a clear pattern
in the wine quality ratings.

#### Were there any interesting or surprising interactions between features?

The interactions between these variables are expected and can be reasoned.
For me, the lack of relationship between residual sugar content and quality
rating was surprising. As residual sugar is strongly correlated to density 
and density is moderately correlated to quality rating, I would have expected
residual sugar to have an effect on quality. 

------

## Final Plots and Summary

#### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
p12 <- ggplot(aes(x = quality, y = alcohol), data = white_wine_data) +
  geom_boxplot(aes(group = quality, fill = factor(quality)), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  ylab("Alcohol (% by volume)") +
  xlab("Quality (score between 0 and 10)") +
  ggtitle("Quality vs. Alcohol") +
  theme(axis.title.x  = element_text(size = 9))

p13 <- ggplot(aes(x = quality, y = density), data = white_wine_data) +
  geom_boxplot(aes(group = quality, fill = factor(quality)), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  ylim(min(white_wine_data$density), 1.0025) +
  ylab("Density (g / cm^3)") +
  xlab("Quality (score between 0 and 10)") +
  ggtitle("Quality vs. Density") +
  theme(axis.title.x  = element_text(size = 9))

p16 <- ggplot(aes(x = quality, y = chlorides), data = white_wine_data) +
  geom_boxplot(aes(group = quality, fill = factor(quality)), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  ylim(min(white_wine_data$chlorides), 0.09) +
  ggtitle("Quality vs. Chlorides") +
  ylab("Chlorides (sodium chloride - g / dm^3)") +
  xlab("Quality (score between 0 and 10)") +
  ggtitle("Quality vs. Chlorides") +
  theme(axis.title.x  = element_text(size = 9))

grid.arrange(p12, p13, p16, ncol = 3)
```

Correlation between Quality and Alcohol:

```{r echo=FALSE}
cor(white_wine_data$quality, white_wine_data$alcohol)
```

Correlation between Quality and Density:

```{r echo=FALSE}
cor(white_wine_data$quality, white_wine_data$density)
```

Correlation between Quality and Chlorides:

```{r echo=FALSE}
cor(white_wine_data$quality, white_wine_data$chlorides)
```


#### Description One

The three variables plotted above side by side, alcohol, density, and chlorides
are the three most correlated variables to quality ratings. While alcohol has a
positive correlation of 0.4356, density has a negative correlation of 0.3012, 
chlorides have a negative correlation of 0.2099.
These variables still only count as moderate/low correlations. Thus, a perfect
linear relationship would not be present. We can see the red line representing 
this trend on the plots.

#### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
p14 <- ggplot(aes(x = density, y = residual.sugar), 
       data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$residual.sugar), 
       quantile(white_wine_data$residual.sugar, 0.95)) +
  xlim(min(white_wine_data$density), quantile(white_wine_data$density, 0.95)) +
  xlab("Density (g / cm^3)") +
  ylab("Residual Sugar (g / dm^3)") +
  scale_color_brewer(type = 'seq', palette = 'OrRd')

p15 <- ggplot(aes(x = density, y = alcohol), data = white_wine_data) +
  geom_point(aes(color = factor(quality)), position = 'jitter') +
  ylim(min(white_wine_data$alcohol), 
       quantile(white_wine_data$alcohol, 0.90)) +
  xlim(min(white_wine_data$density), 
       quantile(white_wine_data$density, 0.95)) +
  xlab("Density (g / cm^3)") +
  ylab("Alcohol (% by volume)") +
  scale_color_brewer(type = 'seq', palette = 'OrRd')

grid.arrange(p14, p15, ncol = 1, top = 'Overview of the Two Strongest Correlations, Colored by Quality')
```

Correlation between Residual Sugar and Density:

```{r echo=FALSE}
cor(white_wine_data$residual.sugar, white_wine_data$density)
```

Correlation of Alcohol and Density:

```{r echo=FALSE}
cor(white_wine_data$alcohol, white_wine_data$density)
```

#### Description Two

There are two strong correlations in the dataset that can have an effect 
on our feature of interest, the quality rating; so only those two relationships
are visualized above. As we can see on the first graph, there is a positive 
correlation of 0.839 between residual sugar and denstity. Also, from the second
graph we can state that there is a negative correlation of 0.7801 between 
alcohol and density.  

Since we have already noticed that there is a moderate negative correlation 
between density and quality, it is not surprising that we can see a trend from 
the graph. 

From the second graph, a different kind of pattern can be described. We can 
clearly notice that as the alcohol content increases, so does the quality rating.
As we know, alcohol and quality rating have a moderately positive correlation 
and as such, quality increases with increasing alcohol levels.  We also noted 
that density has a moderately negative correlation with quality.

#### Plot Three

Summary of Correlations:

```{r echo=FALSE, Plot_Three}
x %>% 
  mutate(rowname = factor(rowname, levels = rowname[order(quality)])) %>%  
  # Ordering by correlation strength
  ggplot(aes(x = rowname, y = quality)) +
    geom_bar(stat = "identity") +
    ylab("Correlation with Quality Ratings") +
    xlab("Variable") +
    theme(axis.text.x  = element_text(angle = 90, vjust = 0.5, size = 12),
          plot.title = element_text(hjust = 0.5, size = 16)) +
    ggtitle("Correlations of Variables with Quality Ratings")
```

#### Description Three

Overall, this last graph sums up the relationship of our variable of interest
to all other variables in the dataset the best. The strongest negative 
relationships in order of strength are density, chorides, volatile acidity, 
total sulfur dioxide, fixed acidity, and residual sugar. The strongest positive
relationships are alcohol, pH, and sulphates.    
Based on the correlations, free sulfur dioxide and citric acid are the least
important determinants of quality.

------

## Reflection

In this dataset of almost 5000 white wine samples from a Portugese region, we
were able to determine that alcohol content, density, and chlorides are the
most important variables to affect the quality ratings.

There are a number of limitations, however, to this dataset which can affect the
outcomes of this analysis. The collected data consists of samples only from a 
specific region in Portugal. Thus, the results may not be accurate for 
representing the general population of all wines. One recommendations for future
analysis would be to investigate wine from different regions and compare the 
results in order to draw valid conclusions on the effects of ingredients and
wine quality. Another limitation is that personal taste preferences are not 
necessarily objective and quality ratings were given by people who might have
favored one wine to the other, so I would also suggest repeating this study
with different people but the same wine samples to judge the quality.  


## References

Dataset obtained from:
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib
